Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-03-27 paper/paperDraft.tex
Add intro

\section{Introduction}
\label{sec:intro}
%- Du bruit au soundscape
%- La pleasantness premier axe sur les affects liés au soundscape
%- On a des modèles de pleasantness perceptifs qui font intervenir des temps de présence des sources
%- Peut-on à partir de la mesure physique faire des modèles de pleasantness sur la base d'indicateurs physiques?
%- Idée: si l'on fait de la reconnaissance de sources, on peut faire un meilleur modèle physique que les précédents
%- Première étape: On vérifie que le corpus créé est proche d'un corpus réel
%- Sur le corpus on fait de l'apprentissage profond
%- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)

@@

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project \ml{a citer si pas fait avant}, and 75 simulated scenes.

@@

\ml{ Je prendrais un point de vue plus general, tout le monde fait alors on utilise ca comme ca ca montre la capacité de generalisation de l'approche // The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}. Thus, a prediction model should take as input the same representation.} Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments. \ml{les resultats seront discutés ensuite ou non ? Sinon le dire ici}

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by LeakyReLU \ml{un peu technique, utiliser les terms complets et citer} activations. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\

@@

Next, the capacity of these predictions to give relevant pleasantness estimates is studied. For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. \ml{Je ne comprends pas : First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data.} This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

The perceptual models of pleasantness \ml{models} are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
\end{align}
\ml{Enlever ou reformuler: Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result.} First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\

@@

Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Since the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered as "weak" labels. Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\ml{faut faire attention au wording, je n'utiliserai pas bon et mauvais, mais plus le fait que les labels sont plus ou moins haut niveau ?}\\

>>>>

\section{Introduction}
\label{sec:intro}
%- Du bruit au soundscape
%- La pleasantness premier axe sur les affects liés au soundscape
%- On a des modèles de pleasantness perceptifs qui font intervenir des temps de présence des sources
%- Peut-on à partir de la mesure physique faire des modèles de pleasantness sur la base d'indicateurs physiques?
%- Idée: si l'on fait de la reconnaissance de sources, on peut faire un meilleur modèle physique que les précédents
%- Première étape: On vérifie que le corpus créé est proche d'un corpus réel
%- Sur le corpus on fait de l'apprentissage profond
%- On compare le nouveau modèle physique deep en comparaison aux modèles physiques old school Cartazur (dimension purement niveau sonore) et Grafic (niveau sonore + Fréquence/sources)
In urban areas, the decreasing quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1 (CITE) norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied (CITE). Specifically, the axis of pleasantness is increasingly associated with soundscape quality in the recent literature (CITE). Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments (CITE). While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness: traffic, human voices and birds. In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\

The perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the construction of perceptually motivated noise maps. With the advent of the Internet of Things and the availability of low-cost sensor networks (CITE), predicting pleasantness from acoustic measurements is an interesting and ongoing area of research as an alternative to systematic human quality assessment. Ricciardi et al. (CITE) present a physical model of pleasantness using only variations of the measured sound level. Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments. Aumond et al. (CITE) explore new indicators based on spectral variations in recordings of sound scenes. These indicators are designed to represent the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. Following the recent advances in machine learning, it may be possible to better approximate the perception of sound sources contributing to pleasantness by developing new indicators relying on source recognition models. However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings. To our knowledge such databases do not currently exist in the literature. A possible alternative is the use of simulated corpora (CITE) where sound scenes are composed from semi-randomly generated annotations.\\

The aim of this research is two-fold. First, a perceptual test is conducted where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction. Next, a deep learning architecture is designed and trained that performs multi-label sound source recognition. Estimations of pleasantness using these predictions are compared to existing models from physical indicators.

@@

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.

@@

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$ . The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\

@@

Next, the capacity of these predictions to give relevant pleasantness estimates is studied. For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. First, pleasantness estimations are equally performant using the model's predictions or the ground truth labels , with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

The perceptual models of pleasantness are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
\end{align}
Several hypotheses can be drawn from this result. First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\

@@

Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels. Thus, in some cases the deep learning model's predictions are considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\
